---
title: "Wildfire Prediction Analysis"
output:
  html_document: default
  html_notebook: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```


## Libraries {#libraries}

```{r libraries}

library(RSQLite)
library(dbplyr)
library(tidyverse)
library(caret)
library(rpart.plot)
library(knitr)
library(kableExtra)
library(e1071)
```

## Getting the Data {#get_data}


```{r load_data}

# create db connection
conn <- dbConnect(SQLite(), 'FPA_FOD_20170508.sqlite')


# pull the fires table into RAM
fires <- tbl(conn, "Fires") %>% collect()

# disconnect from db
dbDisconnect(conn)

```

```{r, echo = FALSE}

# check size
print("Size of data in RAM:")
print(object.size(fires), units = 'Gb')

```

```{r}

print(dim(fires))

```

The dataframe is a little under 1Gb in size and has 1.88 million rows. 

## Sampling the Data {#sample}


```{r sample_data}

# Randomly sample observations from the data
set.seed(123)
index <- sample(c(TRUE, FALSE), nrow(fires), replace = TRUE, prob = c(0.1, 0.9))
fires <- fires[index, ]

```

## Target Feature Analysis {#target}



```{r plot_classes}

fires %>% group_by(STAT_CAUSE_DESCR) %>%
    summarize(n = n()) %>%
    ggplot(aes(x = reorder(STAT_CAUSE_DESCR, -n), y = n/1000)) + 
    geom_bar(stat = 'identity', fill = 'orange') + 
    labs(x = '', y = '# of instances (thousands)', title = 'Target class counts') + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

```



## Data Setup {#data_setup}



```{r train_test_split}

# features to use
features <- c('FIRE_SIZE')

fires$STAT_CAUSE_DESCR <- as.factor(fires$STAT_CAUSE_DESCR)

# index for train/test split
set.seed(123)
train_index <- sample(c(TRUE, FALSE), nrow(fires), replace = TRUE, prob = c(0.8, 0.2))
test_index <- !train_index

# Create x/y, train/test data
x_train <- as.data.frame(fires[train_index, features])
y_train <- fires$STAT_CAUSE_DESCR[train_index]

x_test <- as.data.frame(fires[test_index, features])
y_test <- fires$STAT_CAUSE_DESCR[test_index]


```

## Iteration 1: Benchmark {#benchmark}



```{r benchmark}

preds <- rep('Debris Burning', length(y_test))

test_set_acc <- round(sum(y_test == preds)/length(preds), 4)
print(paste(c("Accuracy:" , test_set_acc)))

```



## Iteration 2: A Simple Decision Tree {#simple}


```{r, tree_1}

# create the training control object.
tr_control <- trainControl(method = 'cv', number = 3)

# Train the decision tree model
set.seed(123)
dtree <- train(x = x_train,
               y = y_train,
               method = 'rpart',
               trControl = tr_control)

```

```{r predictions_1}

# make predictions using test set
preds <- predict(dtree, newdata = x_test)

# calculate accuracy on test set
test_set_acc <- round(sum(y_test == preds)/length(preds), 4)
print(paste(c("Accuracy:" , test_set_acc)))

```


```{r tree_resample_1}

print(dtree$resample)

```


### Results {.tabset}

#### Confusion Matrix Heatmap

```{r conf_heatmap_1}

confusionMatrix(y_test, preds)$table %>%
    prop.table(margin = 1) %>%
    as.data.frame.matrix() %>%
    rownames_to_column(var = 'actual') %>%
    gather(key = 'prediction', value = 'freq',-actual) %>%
    ggplot(aes(x = actual, y = prediction, fill = freq)) +
    geom_tile() +
    geom_text(aes(label = round(freq, 2)), size = 3, color = 'gray20') + 
    scale_fill_gradient(high = 'Red', low = 'Yellow', limits = c(0,1), name = 'Relative Frequency') +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
    ggtitle('Confusion Matrix - Simple Decision Tree')
    

```


#### Confusion Matrix Table

```{r conf_table_1}

# show confusion matrix
confusionMatrix(y_test, preds)$table %>%
    as.data.frame.matrix() %>%
    kable("html") %>%
    kable_styling(bootstrap_options = c('striped'), font_size = 8) %>%
    scroll_box(height = "400px")

```


#### Tree Plot



```{r tree_plot_1}

rpart.plot(dtree$finalModel)

```


#### CV Plot

```{r}

plot(dtree)

```



## Iteration 3: Moar Features! {#moar}



```{r train_test_split_1}

# features to use
features <- c('FIRE_YEAR', 'FIRE_SIZE', 'DISCOVERY_DOY')

x_train <- as.data.frame(fires[train_index, features])
y_train <- fires$STAT_CAUSE_DESCR[train_index]

x_test <- as.data.frame(fires[test_index, features])
y_test <- fires$STAT_CAUSE_DESCR[test_index]

```


```{r tree_2}

# Train the decision tree model
set.seed(123)

dtree <- train(x = x_train,
               y = y_train,
               method = 'rpart',
               tuneLength = 5,
               trControl = tr_control)

```

```{r predictions_2}

# make predictions using test set
preds <- predict(dtree, newdata = x_test)

# calculate accuracy on test set
test_set_acc <- sum(y_test == preds)/length(preds)
print(paste(c("Accuracy:" , round(test_set_acc, 4))))

```



```{r resample_2}

print(dtree$resample)

```



### Results {.tabset}

#### Confusion Matrix Heatmap

```{r conf_heatmap_2}

confusionMatrix(y_test, preds)$table %>%
    prop.table(margin = 1) %>%
    as.data.frame.matrix() %>%
    rownames_to_column(var = 'actual') %>%
    gather(key = 'prediction', value = 'freq',-actual) %>%
    ggplot(aes(x = actual, y = prediction, fill = freq)) +
    geom_tile() +
    geom_text(aes(label = round(freq, 2)), size = 3, color = 'gray20') + 
    scale_fill_gradient(low = 'yellow', high = 'red', limits = c(0,1), name = 'Relative Frequency') +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    ggtitle('Confusion Matrix - Decision Tree')
    
```

#### Confusion Matrix Table

```{r conf_table_2}
# show confusion matrix
confusionMatrix(y_test, preds)$table %>%
    as.data.frame.matrix() %>%
    kable("html") %>%
    kable_styling(bootstrap_options = c('striped'), font_size = 8) %>%
    scroll_box(height = "400px")

```




#### Tree Plot



```{r tree_plot_2}

rpart.plot(dtree$finalModel)

```



#### CV Plot

```{r}

plot(dtree)

```

## Iteration 4: Even More Features {#even_more}


```{r, train_test_split_3}

# features to use
features <- c('FIRE_YEAR', 'FIRE_SIZE', 'DISCOVERY_DOY', 'LATITUDE', 'LONGITUDE')

x_train <- as.data.frame(fires[train_index, features])
y_train <- fires$STAT_CAUSE_DESCR[train_index]

x_test <- as.data.frame(fires[test_index, features])
y_test <- fires$STAT_CAUSE_DESCR[test_index]

```



```{r tree_3}

# Train the decision tree model
set.seed(123)
dtree <- train(x = x_train,
               y = y_train,
               method = 'rpart',
               tuneLength = 8,
               trControl = tr_control)

```

```{r predictions_3}

# make predictions using test set
preds <- predict(dtree, newdata = x_test)

# calculate accuracy on test set
test_set_acc <- sum(y_test == preds)/length(preds)
print(paste(c("Accuracy:" , round(test_set_acc, 4))))


```



### Results {.tabset}

#### Confusion Matrix Heatmap

```{r conf_heatmap_3}

confusionMatrix(y_test, preds)$table %>%
    prop.table(margin = 1) %>%
    as.data.frame.matrix() %>%
    rownames_to_column(var = 'actual') %>%
    gather(key = 'prediction', value = 'freq',-actual) %>%
    ggplot(aes(x = actual, y = prediction, fill = freq)) +
    geom_tile() + 
    geom_text(aes(label = round(freq, 2)), size = 3, color = 'gray20') + 
    scale_fill_gradient(low = 'yellow', high = 'red', limits = c(0,1), name = 'Relative Frequency') +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
    ggtitle('Confusion Matrix - Decision Tree')
    
```

#### Confusion Matrix Table

```{r conf_table_3}

# show confusion matrix
confusionMatrix(y_test, preds)$table %>%
    as.data.frame.matrix() %>%
    kable("html") %>%
    kable_styling(bootstrap_options = c('striped'), font_size = 8) %>%
    scroll_box(height = "400px")

```


#### Tree Plot



```{r tree_plot_3}

rpart.plot(dtree$finalModel)

```

#### CV Plot

```{r}

plot(dtree)

```


###



## Iteration 5: Ensembling {#ensemble}



```{r random_forest}

# Train the decision tree model
set.seed(123)
rfmodel <- train(x = x_train,
                 y = y_train,
                 method = 'rf',
                 tuneLength = 3,
                 ntree = 100,
                 trControl = tr_control)

```

```{r predictions_4}

# make predictions using test set
preds <- predict(rfmodel, newdata = x_test)

# calculate accuracy on test set
test_set_acc <- sum(y_test == preds)/length(preds)
print(paste(c("Accuracy:" , round(test_set_acc, 4))))

```



```{r resample_3}

print(rfmodel$resample)
```



### Results {.tabset}

#### Confusion Matrix Heatmap

```{r conf_heatmap_4}

confusionMatrix(y_test, preds)$table %>%
    prop.table(margin = 1) %>%
    as.data.frame.matrix() %>%
    rownames_to_column(var = 'actual') %>%
    gather(key = 'prediction', value = 'freq',-actual) %>%
    ggplot(aes(x = actual, y = prediction, fill = freq)) +
    geom_tile() +
    geom_text(aes(label = round(freq, 2)), size = 3, color = 'gray20') + 
    scale_fill_gradient(low = 'yellow', high = 'red', limits = c(0,1), name = 'Relative Frequency') + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    ggtitle('Confusion Matrix - Random Forest (100 Trees)')
    
```

#### Confusion Matrix Table

```{r conf_table_4}

# show confusion matrix
confusionMatrix(y_test, preds)$table %>%
    as.data.frame.matrix() %>%
    kable("html") %>%
      kable_styling(bootstrap_options = c('striped'), font_size = 8) %>%
  scroll_box(height = "400px")

```

#### CV Results

```{r}

plot(rfmodel)

```

### Iteration 6: Boosting {#boosting}



```{r}

fires$BURN_TIME <- fires$CONT_DATE - fires$DISCOVERY_DATE

```


```{r}

# features to use
features <- c('FIRE_YEAR', 'FIRE_SIZE', 'DISCOVERY_DOY', 'LATITUDE', 'LONGITUDE', 'BURN_TIME')

x_train <- as.data.frame(fires[train_index, features])
y_train <- fires$STAT_CAUSE_DESCR[train_index]

x_test <- as.data.frame(fires[test_index, features])
y_test <- fires$STAT_CAUSE_DESCR[test_index]


```


```{r xgboost}

tr_control <- trainControl(
    method = 'cv',
    number = 2,
    verboseIter = FALSE,
    allowParallel = TRUE)
    
tune_grid <- expand.grid(
    nrounds = c(100),
    max_depth = c(8),
    eta = c(0.1),
    gamma = c(0.01),
    colsample_bytree = c(0.75),
    subsample = c(0.5),
    min_child_weight = c(0))

# Train the decision tree model
set.seed(123)
xgbmodel <- train(
    x = x_train,
    y = y_train,
    method = 'xgbTree',
    trControl = tr_control,
    tuneGrid = tune_grid)

```

```{r predictions_5}

# make predictions using test set
preds <- predict(xgbmodel, newdata = x_test)

# calculate accuracy on test set
test_set_acc <- sum(y_test == preds)/length(preds)
print(paste(c("Accuracy:" , round(test_set_acc, 4))))

```


```{r resample_4}

print(xgbmodel$resample)

```


### Results {.tabset}

#### Confusion Matrix Heatmap

```{r conf_heatmap_5}

confusionMatrix(y_test, preds)$table %>%
    prop.table(margin = 1) %>%
    as.data.frame.matrix() %>%
    rownames_to_column(var = 'actual') %>%
    gather(key = 'prediction', value = 'freq',-actual) %>%
    ggplot(aes(x = actual, y = prediction, fill = freq)) +
    geom_tile() +
    geom_text(aes(label = round(freq, 2)), size = 3, color = 'gray20') + 
    scale_fill_gradient(low = 'yellow', high = 'red', limits = c(0,1), name = 'Relative Frequency') + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    ggtitle('Confusion Matrix - xgboost')
    
```

#### Confusion Matrix Table

```{r conf_table_5}

# show confusion matrix
confusionMatrix(y_test, preds)$table %>%
    as.data.frame.matrix() %>%
    kable("html") %>%
      kable_styling(bootstrap_options = c('striped'), font_size = 8) %>%
  scroll_box(height = "400px")

```

#### CV Results

```{r}

#plot(xgbmodel)

```
